---
title: "SAWASDEE_PFAS_Breastmilk"
author: "Neha Sehgal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
# Load Libraries
library('haven')
library('table1')
library('tidyverse')
library('Hmisc')
library('ggplot2')
library('ggcorrplot')
library('corrplot')
library(dplyr)
library(ggpubr)
library("tibble")
library("sjPlot")
library(readxl)
library(naniar)
library(DataExplorer)
library(viridis)
library(stringr)
```

## Introduction

This is the final R code for the main results, figures, and tables
included in the publication: ***Eick et al (2024). Per- and
polyfluoroalkyl substances in serum and breastmilk samples among
pregnant farmworkers in Thailand.*** doi:https://doi.org/10.1016/j.ijheh.2024.114509

## Data Import: Load clean data

This data includes serum and breastmilk PFAS, covariates (for analytical
subset and the parent SAWASDEE cohort) and NHANES data. Please note
covariate data has been cleaned and recoded. Specifically, for this
analysis, PFAS data has transformed so that:

1.  undergone QA/QC,
2.  cleaned to remove any duplicates
3.  single imputed with LOQ/sqrt(2), and
4.  natural-log using R's log() function
5.  sumPFOS = sum of linear and branched PFOS

```{r import, eval=FALSE, include=FALSE}
# #### load PFAS data 
# load("SAWASDEE_PFAS_clean_NS_10282023.Rda")
# 
# #### load covariate data
# load("breastfeed_PFAS_covars.Rda")
# load("PFAS_TE_EDI_covars.Rda")
# load("sawasdee_raw.Rda")
# 
# #### load NHANES data
# load("nhanes_final_dataset_2023.Rda")
```

## Study Demographics 

**Table 1.** Study demographics of the Study of Asian Women and
Offspring’s Development and Environmental Exposures (SAWASDEE) cohort,
2017-2019

```{r demographics, echo=FALSE}
#### Table 1. for analytical susbet 
table1(~ child_sex+mom_education+ppbmi+location+ mat_age+Parity+household_adj_income, data = breastfeed_PFAS_covars)

#### Table 1. for parent SAWASDEE cohort 
table1(~Infant_Sex+Maternal_education+ BMI_delivery + Location+ `Maternal Age` + as.character(Parity) + household_adj_income, sawasdee_raw)
```

## PFAS

**Table 2.** Distribution of serum and breastmilk per- and
poly-fluoroalkyl substances (PFAS) concentrations (ng/mL) in the Study
of Asian Women and Offspring’s Development and Environmental Exposures
(SAWASDEE cohort, 2017-2019) and among reproductive aged women in the
National Health and Nutrition Examination Study NHANES; years
2017-2018).

```{r pfas, message=FALSE, warning=FALSE}
#### distributions using qq plots and histograms
plot_histogram(PFAS)
plot_qq(PFAS, nrow  = 4)

PFAS_fill_visit <- PFAS %>%  mutate(pfas_visit_filled = case_when(pfas_visit  == "10" ~ "9", 
                                                                  pfas_visit == "12" ~ "11", 
                                                                  TRUE ~ pfas_visit))
### create total PFOS ### 
PFAS_fill_visit <- PFAS_fill_visit %>% rowwise() %>%  mutate(sumPFOS = `L-PFOS (ng/mL)` + `B-PFOS-Q`, 
                                                             lnsPFOS =  log(sumPFOS))

# PFAS: use all PFAS  from Stephanie Eick: Eick SAWASDEE PFAS 10-19-2023.xlsx  provided as of Oct 25 2023
# note, dR. PRINN PANUWET @ Rsph, emory says that we only use xx_Q as its the  quantitation ions. No use xxx_c.
PFAS.vars <- c("PFHpA (ng/mL)" , "L-PFOA (ng/mL)" , "PFHxS (ng/mL)" , "PFNA (ng/mL)" , "PFDA (ng/mL)" ,
               #"sumPFOS" ,
               "L-PFOS (ng/mL)", "B-PFOS-Q" , "PFUnDA (ng/mL)" , "PFDoDA (ng/mL)")
#### Table 2. 
# function for summary distribution table
table1.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x), digits=2), c("",
                                                           "Mean (SD)"=sprintf("%s (%s)", MEAN, SD)))}
tabe1.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
                                                  sprintf("%d (%0.0f %%)", FREQ, PCT))))}
table2.quartiles <- function(x, name, data2, ...) {
  Q05 <- quantile(x, 0.05, na.rm = T)
  Q1 <- quantile(x, 0.25, na.rm = T)
  median <- median(x, na.rm = T)
  Q3 <- quantile(x, 0.75, na.rm = T)
  Q95 <- quantile(x, 0.95, na.rm = T)
  
  out <- c("",
           "5%" = sprintf("%.2f", Q05),
           "25%" = sprintf("%.2f", Q1), 
           "50%" = sprintf("%.2f", median), 
           "75%" = sprintf("%.2f", Q3), 
           "95%" = sprintf("%.2f", Q95))
  out}

#### Table 2 overall
table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + sumPFOS + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled*Matrix, 
       data = PFAS_fill_visit, 
       render.continuous=c("Geo. mean (Geo. SD)"="GMEAN (GSD)"))

table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + sumPFOS + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled*Matrix, 
       data = PFAS_fill_visit, 
       render.continuous=table2.quartiles) # get min, median, and max. 

# by matrix 
PFAS.serum.keep<- PFAS %>% filter(Matrix == "Serum") %>%  mutate(pfas_visit_filled = case_when(pfas_visit  == "10" ~ "9",
                                                                                               pfas_visit  == "10M" ~ "9",
                                                                                               pfas_visit  == "9M" ~ "9",
                                                                                               pfas_visit == "12" ~ "11", 
                                                                                               pfas_visit == "12M" ~ "11", 
                                                                                               pfas_visit == "11M" ~ "11", 

                                                                                               TRUE ~ pfas_visit)) %>%  select(ptid, pfas_visit_filled, all_of(PFAS.vars)) 

PFAS.bm.keep<- PFAS %>% filter(Matrix == "Breast milk") %>% mutate(pfas_visit_filled = case_when(pfas_visit  == "10" ~ "9",
                                                                                               pfas_visit  == "10M" ~ "9",
                                                                                               pfas_visit  == "9M" ~ "9",
                                                                                               pfas_visit == "12" ~ "11", 
                                                                                               pfas_visit == "12M" ~ "11", 
                                                                                               pfas_visit == "11M" ~ "11",                                                                                                  TRUE ~ pfas_visit)) %>% select(ptid, pfas_visit_filled, all_of(PFAS.vars)) 
# serum pfas
table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + `sumPFOS` + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled, 
       data = PFAS.serum.keep, 
       render.continuous=c("Geo. mean (Geo. SD)"="GMEAN (GSD)"))
table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + `sumPFOS` + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled, 
       data = PFAS.serum.keep, 
       render.continuous=table2.quartiles) # get min, median, and max. 

# breastmilk pfas 
table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + sumPFOS + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled, 
       data = PFAS.bm.keep, 
       render.continuous=c("Geo. mean (Geo. SD)"="GMEAN (GSD)"))
table1(~`PFHpA (ng/mL)` + `L-PFOA (ng/mL)` + `PFHxS (ng/mL)` + `PFNA (ng/mL)` + `PFDA (ng/mL)` + sumPFOS  + `L-PFOS (ng/mL)`+ `B-PFOS-Q` + `PFUnDA (ng/mL)` + `PFDoDA (ng/mL)`|pfas_visit_filled, 
       data = PFAS.bm.keep, 
       render.continuous=table2.quartiles) # get min, median, and max. 

# percent above LOD  = (Total N - #NAs) / Total N]
calc_above_lod <- function(x) {x  = case_when(x == "<LOQ" ~ "below LOQ",
                                              x == NA ~ NA, 
                                              TRUE ~ "above LOQ")}

pfas <- PFAS %>%  mutate(pfas_visit_filled = case_when(pfas_visit  == "10" ~ "9", 
                                               pfas_visit == "12" ~ "11", 
                                               TRUE ~ pfas_visit))
                         # sumPFOS = `L-PFOS (ng/mL)` + `B-PFOS-Q`, 
                         # lnsPFOS =  log(sumPFOS))

PFAS.serum.loq <- pfas %>%  filter(Matrix == "Serum" & ptid %in% PFAS.serum.keep$ptid) %>%  mutate_at(vars(c(all_of(PFAS.vars))), calc_above_lod)
PFAS.bm.loq <- pfas %>%  filter(Matrix == "Breast milk" & ptid %in% PFAS.bm.keep$ptid) %>%  mutate_at(vars(c(all_of(PFAS.vars))), calc_above_lod)

#### calculate % above lod 
table1(~.|PFAS.bm.loq$pfas_visit_filled, data = PFAS.bm.loq[PFAS.vars])
table1(~.|PFAS.serum.loq$pfas_visit_filled, data = PFAS.serum.loq[PFAS.vars])
```

**Calculate lactational transfer efficiency (TE) and estimated daily
intake (EDI)**

We assessed the lactational transfer of individual PFAS from serum and
breastmilk using transfer efficiencies (TEs, %). For both 9.5 and 11.5
months, we calculated TEs using the equation TE = (C~BM~/C~S~)\*100. In
this equation, C~BM~ is the concentration of a given PFAS in breastmilk
and C~S~ is the concentration of a given PFAS in serum.

```{r te}
#### TE% individually
# save all the raw PFAS data
PFAS_fill_visit_wide_raw <- PFAS_fill_visit  %>%  select(ptid, pfas_visit_filled, Matrix, all_of(PFAS.vars), "sumPFOS") %>%  
  mutate(time = str_c(pfas_visit_filled, Matrix, sep = "_")) %>%  select(ptid, time, all_of(PFAS.vars), "sumPFOS")

PFAS_fill_visit_wide_raw <- PFAS_fill_visit_wide_raw %>% 
  pivot_wider(names_from = time,  
              values_from = c("PFHpA (ng/mL)","L-PFOA (ng/mL)", "PFHxS (ng/mL)","PFNA (ng/mL)", "PFDA (ng/mL)",
                              "L-PFOS (ng/mL)", "B-PFOS-Q","PFUnDA (ng/mL)", "PFDoDA (ng/mL)", "sumPFOS" ))

# for-loop the TE% formula on each PFAS                           
all.PFAS.vars <- c(PFAS.vars, "sumPFOS")
PFAS_TE_individual <- data.frame(PFAS_fill_visit_wide_raw$ptid) 
colnames(PFAS_TE_individual) <- "ptid"

for( i in all.PFAS.vars){
  subset <-  PFAS_fill_visit_wide_raw %>%  select(ptid, starts_with(i))  # subset the 3 measurements for a single PFAS 
  subset$bm9_s5 <-  (subset[,3] / subset[,2])*100                              # calc breastmilk 9mo to serum TE%
  subset$bm11_s5 <- (subset[,4] / subset[,2])*100                              # calc breastmilk 11mo to serum TE%
  
  names(subset) = gsub(pattern = "bm9_s5", replacement = paste0("TE_", i , "_9b_5m"), x = names(subset)) # beautiful names 
  names(subset) = gsub(pattern = "bm11_s5", replacement = paste0("TE_", i , "_11b_5m"), x = names(subset)) # beautiful names
  
  subset <- subset %>%  select(ptid, starts_with("TE"))
  PFAS_TE_individual <- left_join(PFAS_TE_individual, subset, by = "ptid", copy = F)   # join data
}

PFAS_TE_individual <- as.list(PFAS_TE_individual)   # convert list to data frame
PFAS_TE_individual<- do.call(cbind, PFAS_TE_individual)
PFAS_TE_individual <- as.data.frame(PFAS_TE_individual)
names(PFAS_TE_individual) = gsub(pattern = "(_Breast milk)", replacement = paste0("b_5s"), x = names(PFAS_TE_individual)) # beautiful names
colnames(PFAS_TE_individual)[2:21] <- paste('TE', colnames(PFAS_TE_individual)[2:21], sep = '_') # ADD  a prefix of TE that denotes TE%

#### summarize 
table1(~., data = PFAS_TE_individual[2:21])
```

We calculated the estimated daily intake (EDI) of individual PFAS
occurring via breastfeeding. EDIs were calculated using the equation
EDI= concentration of a given PFAS in breastmilk\*food ingestion rate.
For this equation, the food ingestion rate (mL/bodyweight in kilograms
[kg]/day), which represents the average daily intake of breastmilk, was
assumed to be 150 mL/kg/day for 6-12 month old infants.

```{r edi}
####  calculate EDI for each individual#### 
PFAS_EDI_ind <-PFAS_fill_visit_wide_raw %>%  select(ptid, ends_with("Breast milk"))
PFAS_EDI_ind<- PFAS_EDI_ind %>%  mutate_at(vars(2:21), .funs= funs(.*150))
# summary 
table1(~., data = PFAS_EDI_ind[2:21])

PFAS_EDI_ind_long <- PFAS_EDI_ind %>%  pivot_longer( cols = 2:21,
                                                     names_to='EDI_visit',
                                                     values_to='EDI')

PFAS_EDI_ind_long <- PFAS_EDI_ind_long %>% mutate(pfas = str_split_i(EDI_visit, "_", -3),
                                                  EDI_visit2 = str_split_i(EDI_visit, "_", -2),
                                                  EDI_visit2 = paste0(EDI_visit2, "mo Breastmilk"))

#### summarize
table1(~`EDI`|pfas+EDI_visit2, data = PFAS_EDI_ind_long[2:5])
```

**Correlation between TE and EDI**

**Figure 3.** Spearman correlation between natural log transformed per-
and poly-fluoroalkyl substances (PFAS) measured at different time points
pre- and post-delivery that had \>50% detection levels in the Study of
Asian Women and Offspring’s Development and Environmental Exposures
(SAWASDEE cohort, 2017-2019).

```{r corrplot}
#### data arrangenent for corelation plot 
corr_data<- PFAS_fill_visit  %>%  select(starts_with("ln"), ptid, pfas_visit_filled, Matrix) %>%  
  mutate(time = str_c(pfas_visit_filled, Matrix, sep = "_")) %>%  
  select(time, ptid, 
         "lnPFHpA", "lnLPFOA", "lnPFHxS" , "lnPFNA",  "lnPFDA" , 
         "lnPFUnDA" ,   "lnPFDoDA", "lnsPFOS" )

corr_data <- corr_data %>%  pivot_wider(names_from = time, 
                                        values_from = c("lnPFHpA", "lnLPFOA", "lnPFHxS" , "lnPFNA",  "lnPFDA" , 
                                                        "lnPFUnDA", "lnPFDoDA", "lnsPFOS" ))

corr_data <- corr_data %>% mutate_all(as.numeric)
corr_data <- corr_data %>% select(ptid ,lnLPFOA_5_Serum,
                                  lnPFHxS_5_Serum, 
                                  lnPFNA_5_Serum, 
                                  lnPFDA_5_Serum, 
                                  lnPFUnDA_5_Serum, 
                                  lnsPFOS_5_Serum, 
                                  "lnPFNA_9_Breast milk", 
                                  "lnsPFOS_9_Breast milk", 
                                  "lnsPFOS_11_Breast milk")

names(corr_data) <- c("ptid" , 
                      "PFOA in 2nd Trimester Serum",
                      "PFHxS in 2nd Trimester Serum", 
                      "PFNA in 2nd Trimester Serum", 
                      "PFDA in 2nd Trimester Serum", 
                      "PFUnDA in 2nd Trimester Serum", 
                      "PFOS in 2nd Trimester Serum", 
                      "PFNA in 9.5month Breastmilk", 
                      "PFOS in 9.5month Breastmilk", 
                      "PFOS in 11.5month Breastmilk")

# Function to calculate p-values
cor.mtest <- function(mat, ...) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p.mat <- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

# Calculate correlation matrix and p-values
corr_data <- as.data.frame(corr_data[-1])
cor_matrix <- cor(corr_data, method = "spearman", use = "complete.obs")
p_matrix <- cor.mtest(corr_data, method = "spearman")


#### Plot with p-value stars
corrplot(cor_matrix, method = "color", addCoef.col = 'black', number.digits = 2, 
         title = "", tl.col = "black", type = "lower", 
         number.cex = 0.75, cl.pos = 'n', col = COL2('RdYlBu'), 
         p.mat = p_matrix, sig.level = c(0.05), insig = "label_sig", pch.cex = 1.2, pch.col = "black")

trace(corrplot, edit=TRUE)
# replace line 443 place_points (...)
# adjust text(X,Y ...) according to your needs, here +0.25 is added to the Y-position    
  place_points = function(sig.locs, point) {
   text(pos.pNew[, 1][sig.locs], (pos.pNew[, 2][sig.locs])+0.25,
        labels = point, col = pch.col, cex = pch.cex, lwd = 2)}
```

**Differences in TE and EDI by demographics**

**Table 3.** Distribution of the mean transfer efficiency and
lactational estimated daily intake, overall and by study demographics in
the Study of Asian Women and Offspring’s Development and Environmental
Exposures (SAWASDEE cohort, 2017-2019).  

```{r te_edi_demographics}
###### Data Merging
# join te and breastfeed_pfas_covars
tePFAS_covars<- left_join(breastfeed_PFAS_covars, PFAS_TE_individual, by = "ptid")

# add edi 
tePFAS_covars <- left_join(tePFAS_covars, PFAS_EDI_ind, by = "ptid")
colnames(tePFAS_covars) <- str_replace(colnames(tePFAS_covars), ".y", "_EDI")


####### What Variables are needed
#11 month PFNA, and sum PFOS at both 9 and 11 month timepoints.
# timepoint 
select_pfas_te_edi <- c("TE_PFNA (ng/mL)_9b_5s",
                        "TE_sumPFOS_9b_5s", "TE_sumPFOS_11b_5s", 
                        "PFNA (ng/mL)_9_Breast milk_EDI",
                        "PFNA (ng/mL)_11_Breast milk_EDI", 
                        "sumPFOS_9_Breast milk_EDI", 
                        "sumPFOS_11_Breast milk_EDI")


##### Save results ######
# save(tePFAS_covars, file = "PFAS_TE_EDI_covars.Rda")
########## 

######## Table 3- overall summary
table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s, data = tePFAS_covars)

table1(~ `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` +  `sumPFOS_11_Breast milk_EDI`, data = tePFAS_covars)

######## T-tests/ANOVAs for TE by demographics
pvalue <- function(x, ...) {
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) {
    # For numeric variables, perform a standard 2-sample t-test
    p <- t.test(y ~ g)$p.value
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

# sex
table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s +
       `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI`| child_sex, data = tePFAS_covars, overall=F, extra.col=list(`P-value`=pvalue))

# location 
table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s + 
         `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI`| location, data = tePFAS_covars, overall=F, extra.col=list(`P-value`=pvalue))


# matenral age
# dicotomize at median 
summary(tePFAS_covars$mat_age)
tePFAS_covars$mat_age_cat <- ifelse(tePFAS_covars$mat_age < 25, "below 25 yrs", "25 or above")

table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s + 
         `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI`| mat_age_cat, data = tePFAS_covars, overall=F, extra.col=list(`P-value`=pvalue))

# pre-pregnancy BMI 
tePFAS_covars <- tePFAS_covars %>%  mutate(ppbmi_cat = case_when(ppbmi < 18.5 ~ "<25", 
                                                                 18.5 <= ppbmi &  ppbmi < 24.9	~ "<25",
                                                                 25 <= ppbmi & ppbmi < 29.9	~ "25-29.9", 
                                                                 ppbmi >=30 ~ 	">30", 
                                                                 TRUE ~ "bad") )

table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s + 
         `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI` 
       | ppbmi_cat, data = tePFAS_covars, 
       overall=F)

res <- tePFAS_covars %>% 
  select(all_of(select_pfas_te_edi)) %>%
  map_df(~ broom::tidy(oneway.test(. ~ tePFAS_covars$ppbmi_cat)), .id = 'var')

# Household adj income 
hist(tePFAS_covars$household_adj_income)
tePFAS_covars <-tePFAS_covars %>% mutate(household_adj_income_cat = case_when(household_adj_income <= 1600 | household_adj_income == 1600 ~ "< or below 1600", 
                                                                              household_adj_income > 1600 ~ "> 1600", 
                                                                              TRUE ~ NA)) 

table1(~ `TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s + 
         `PFNA (ng/mL)_9_Breast milk_EDI` +  `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI`
       | household_adj_income_cat, data = tePFAS_covars %>%  drop_na(household_adj_income_cat), 
       overall=T)

res <- tePFAS_covars %>% 
  select(all_of(select_pfas_te_edi)) %>% 
  map_df(~ broom::tidy(t.test(. ~ tePFAS_covars$household_adj_income_cat)), .id = 'var')


# maternal education 
tePFAS_covars <- tePFAS_covars %>%  mutate(mom_education = case_when(mom_education == "College graduate or more" ~ "Junior High school and more", 
                                                                     mom_education == "Diploma/technical school/equivalent" ~ "Junior High school and more", 
                                                                     mom_education == "high school/did not graduate" ~ "Junior High school and more",
                                                                     mom_education == "junior high/high school" ~ "Junior High school and more",
                                                                     mom_education == "None" ~ "Primary school or never attended",
                                                                     mom_education == "primary" ~ "Primary school or never attended") )

table1(~`TE_PFNA (ng/mL)_9b_5s` + TE_sumPFOS_9b_5s +TE_sumPFOS_11b_5s + 
         `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` + `sumPFOS_11_Breast milk_EDI` | mom_education, tePFAS_covars %>%  drop_na(mom_education) )

res <- tePFAS_covars %>% 
  select(all_of(select_pfas_te_edi)) %>% 
  map_df(~ broom::tidy(t.test(. ~ tePFAS_covars$mom_education)), .id = 'var')


# parity
table1(~ `TE_PFNA (ng/mL)_11b_5s` + TE_sumPFOS_9b_5s + TE_sumPFOS_11b_5s | Pari_EDI, data = tePFAS_covars, overall=F, extra.col=list(`P-value`=pvalue))
table1(~ `PFNA (ng/mL)_9_Breast milk_EDI` + `sumPFOS_9_Breast milk_EDI` +  `sumPFOS_11_Breast milk_EDI` | Pari_EDI, data = tePFAS_covars, overall=F, extra.col=list(`P-value`=pvalue))

res <- tePFAS_covars %>% 
  select(all_of(select_pfas_te_edi)) %>%
  map_df(~ broom::tidy(t.test(. ~ tePFAS_covars$Pari_EDI)), .id = 'var')
```

## NHANES comparison

For those PFAS that were also measured in the US National Health and
Nutrition Examination Study (NHANES, years 2017-2018), we compared serum
PFAS levels in the SAWASDEE cohort with reproductive aged women (18-40
years) to facilitate quantitative comparisons with other populations of
reproductive aged women. \

This code produced: 
**Fig. 1.** Violin plots showing the distribution of all measured concentrations (ng/mL; natural log transformed) per- and poly-fluoroalkyl substances (PFAS) in the 
Study of Asian Women and Offspring’s Development and Environmental Exposures (SAWASDEE cohort, 2017–2019) and among reproductive aged women in the 
National Health and Nutrition Examination Study (NHANES; years 2017–2018). 
Abbreviations: ln, natural log.

**Table 2** Distribution of serum and breastmilk per- and poly-fluoroalkyl substances (PFAS) concentrations (ng/mL) in the Study of Asian Women and Offspring’s Development and Environmental Exposures (SAWASDEE cohort, 2017–2019) and among reproductive aged women in the National Health and Nutrition Examination Study (NHANES; years 2017–2018).
```{r nhanes}
#### Cleaning NHANES data and selection
# summary 
summary(nhanes_final_dataset_2023$RIDAGEYR)

# nhanes_final_dataset_2023 <- nhanes_final_dataset_2023 %>% filter(RIDAGEYR < 41 & RIAGENDR == "2") 
# 
# nhanes_final_dataset_2023 <- nhanes_final_dataset_2023 %>%  mutate_at(.vars = 2:11, .funs = as.character)
# 
# nhanes_final_dataset_2023 <- nhanes_final_dataset_2023 %>%  mutate_at(.vars = 12:19, .funs = log) # comment this in only for violin plots, Table1 will not give geo mean or SD with this in. 
# 
# nhanes_final_dataset_2023 <- nhanes_final_dataset_2023 %>%  drop_na()
# 
# nhanes_final_dataset_2023 <- nhanes_final_dataset_2023 %>%  mutate_at(.vars = 12:19, .funs = as.numeric) 


#### Table 2 - NHANES summary
table1(~
         `LBXMFOS`+
         `LBXMPAH`+
         `LBXNFOA`+
         `LBXNFOS`+
         `LBXPFDE`+
         `LBXPFHS`+
         `LBXPFNA`+
         `LBXPFUA`,  data = nhanes_final_dataset_2023,
       render.continuous=c("Geo. mean (Geo. SD)"="GMEAN (GSD)"))

table1(~
         `LBXMFOS`+
         `LBXMPAH`+
         `LBXNFOA`+
         `LBXNFOS`+
         `LBXPFDE`+
         `LBXPFHS`+
         `LBXPFNA`+
         `LBXPFUA`,  data = nhanes_final_dataset_2023, render.continuous=table2.quartiles) 

# lod 
table1(~.,  data = nhanes_final_dataset_2023[2:11], 
       render.continuous=c(.="Mean (CV%)", .="Median [Min, Max]",
                           "Geo. mean (Geo. CV%)"="GMEAN (GCV%)"))

table1(~LBDMFOSL,  data = nhanes_final_dataset_2023, 
       render.continuous=c(.="Mean (CV%)", .="Median [Min, Max]",
                           "Geo. mean (Geo. CV%)"="GMEAN (GCV%)"))

# pivot for violin plot
nhanes_final_dataset_2023_long <- nhanes_final_dataset_2023[c(1, 12:19)] 

nhanes_final_dataset_2023_long<- pivot_longer(nhanes_final_dataset_2023_long, cols = 2:9)


# label 
nhanes_final_dataset_2023_long <- nhanes_final_dataset_2023_long %>%  mutate(name_neat = case_when(
  name == "LBXMFOS" ~ "Sm-PFOS PFOS (ng/mL)", 
  name == "LBXMPAH" ~ "Me-PFOSA-AcOH (ng/mL)", 
  name == "LBXNFOA" ~ "PFOA (ng/mL)", 
  name == "LBXNFOS" ~ "PFOS (ng/mL)", 
  name == "LBXPFDE" ~ "PFDA (ng/mL)", 
  name == "LBXPFHS" ~ "PFHxS (ng/mL)",
  name == "LBXPFNA" ~	"PFNA (ng/mL)", 
  name == "LBXPFUA" ~ "PFUA (ng/mL)"), 
  visit_neat = "2017-2018", 
  study = "NHANES")

##### Data Merging: SAWASDEE and NHANES data and selection
# covert sawasdee data to long data 
PFAS_fill_visit_5serum_BM <- PFAS_fill_visit %>%  select(starts_with("ln"), pfas_visit_filled)
PFAS_fill_visit_5serum_BM <- PFAS_fill_visit_5serum_BM %>%  pivot_longer(cols = 1:10)
PFAS_fill_visit_5serum_BM <-  PFAS_fill_visit_5serum_BM %>%  mutate( name_neat = name, 
                                                                     visit_neat = case_when(pfas_visit_filled == "5" ~ "2nd Trimester Serum",
                                                                                            pfas_visit_filled == "9" ~ "9.5 months Breastmilk",
                                                                                            pfas_visit_filled == "11" ~ "11.5 months Breastmilk"
                                                                     ), 
                                                                     study = "SAWASDEE")

# MERGE SAWASDEE AND nhanes 
pfas_nahanes_sawasdee <- rbind.data.frame(PFAS_fill_visit_5serum_BM[2:6], nhanes_final_dataset_2023_long[2:6])

pfas_nahanes_sawasdee <- pfas_nahanes_sawasdee %>%  mutate(name_neat2 = case_when(
  name_neat == "Sm-PFOS PFOS (ng/mL)" ~ "lnsPFOS", 
  name_neat == "Me-PFOSA-AcOH (ng/mL)" ~ "lnMe-PFOSA-AcOH (ng/mL)", 
  name_neat == "PFOA (ng/mL)" ~ "lnLPFOA", 
  name_neat == "PFOS (ng/mL)" ~ "linear pfos", 
  name_neat == "PFDA (ng/mL)" ~ "lnPFDA", 
  name_neat == "PFHxS (ng/mL)" ~ "lnPFHxS",
  name_neat == "PFNA (ng/mL)"~ "lnPFNA", 
  name_neat == "PFUA (ng/mL)" ~ "lnPFUnDA", 
  TRUE ~ name_neat))

pfas_nahanes_sawasdee <- pfas_nahanes_sawasdee %>%  filter(name_neat2 != "lnBPFOS_q" &
                                                             name_neat2 != "lnLPFOS_c" &
                                                             name_neat2 != "lnLPFOS"  &
                                                             name_neat2 !=  "linear pfos" & 
                                                             name_neat2 != "lnMe-PFOSA-AcOH (ng/mL)" & 
                                                             name_neat2  != "lnPFDoDA" & 
                                                             name_neat2 != "lnPFHpA")

pfas_nahanes_sawasdee <- pfas_nahanes_sawasdee %>%  mutate(visit_neat = ifelse(visit_neat == "2017-2018",  "Serum",visit_neat
))


##### Figure 1: # violin plot to Compare NHANES data and selection
pfas_nahanes_sawasdee$name_neat3 = factor(pfas_nahanes_sawasdee$name_neat2, 
                                          levels=c("lnPFNA","lnLPFOA","lnsPFOS", 
                                                   "lnPFHxS", "lnPFDA", "lnPFUnDA")) 

pfas_nahanes_sawasdee = pfas_nahanes_sawasdee %>%  mutate(name_neat4 = case_when(name_neat3 == "lnPFNA" ~ "PFNA",
                                                                                 name_neat3 == "lnLPFOA" ~ "PFOA",
                                                                                 name_neat3 == "lnsPFOS" ~ "PFOS",
                                                                                 name_neat3 == "lnPFHxS" ~ "PFHxS",
                                                                                 name_neat3 == "lnPFDA" ~   "PFDA",
                                                                                 name_neat3 == "lnPFUnDA" ~ "PFUnDA")) 
pfas_nahanes_sawasdee$name_neat4 = factor(pfas_nahanes_sawasdee$name_neat4, 
                                          levels=c("PFNA","PFOA","PFOS", 
                                                   "PFHxS", "PFDA", "PFUnDA")) 


ggplot(pfas_nahanes_sawasdee, aes(x= factor(visit_neat, level = c("Serum", "2nd Trimester Serum",
                                                                  "9.5 months Breastmilk",
                                                                  "11.5 months Breastmilk" ) ),
                                  y = value, 
                                  fill = visit_neat, color= visit_neat)) + 
  geom_violin(trim = F, alpha = 0.5) +
  geom_jitter(alpha = 0.35, size = 1.2) + 
  ggthemes::scale_fill_tableau(palette = "Tableau 10") +
  ggthemes::scale_color_tableau(palette = "Tableau 10") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) + 
  labs(y = "PFAS (ng/mL) measurements", x = "") + 
  ggtitle("Distribution of PFAS in NHANES vs SAWADEE") + 
  facet_wrap(~name_neat4, ncol = 3, scales = "free") + 
  theme_bw() + 
  theme(axis.title = element_text(face = "bold", size = 12),
        axis.text = element_text(size = 11),
        strip.text = element_text(size = 12),
        legend.position = "none")

ggsave("sawasdee_nhanes_pfas.jpeg")
```

Note: **Figure 2.** Composition profile of per- and poly-fluoroalkyl
substances (PFAS) in the Study of Asian Women and Offspring’s
Development and Environmental Exposures (SAWASDEE cohort, 2017-2019) and
among reproductive aged women in the National Health and Nutrition
Examination Study (NHANES; years 2017-2018 ] was made using SAS code.

##### End of code
